---
title: 'Data Analysis Project'
author: "STAT 420, Fall 2024, C. Easton, A. Roh, T. Toter, A. Treptow"
date: 'December xx, 2024'
output:
  pdf_document:
    toc: yes
  html_document:
    theme: readable
    toc: yes
urlcolor: cyan
---

**Introduction:**

In this data analysis study we appled the principles we learned during the semester to iterate on a linear regression model.

For our case study, we chose a "Wine Quality" dataset from the UC Irvine Machine Learning Repository. The data relates to red and white variants of the Portuguese vinho verde wine samples. We drew the data from the following site:

  https://archive.ics.uci.edu/dataset/186/wine+quality

Each wine sample contains a record of 10 numerically measured physicochemical attributes, such as acidity, residual sugar, chlorides, and pH. We combined two datasets (one for white wine and one for red wine), resulting in an additional categorical variable for wine type.

The 11 attributes listed above served as our source predictor variables. A final attribute, measuring quality, served as our response variable. Each quality measurement is a subjectively assigned integer, ranging from 1 to 10. Our goal was to build a model that could use the objectively measured predictors as inputs to estimate how a human would rate the wine.

```{r setup}

options(repos = c(CRAN = "https://cloud.r-project.org"))

```

## load data

```{r}

# Load Necessary Libraries
library(dplyr) # For data manipulation

# Load the Red Wine Data
red_wine = read.csv("winequality-red.csv", sep = ";")

# Load the White Wine Data
white_wine = read.csv("winequality-white.csv", sep = ";")

# Add a Categorical Variable for Wine Type
red_wine$type = "Red"
white_wine$type = "White"

# Combine the Two Datasets
wine_data = bind_rows(red_wine, white_wine)


# Convert 'type' to a Factor
wine_data$type = as.factor(wine_data$type)

# Categorize the 'alcohol' Predictor
#wine_data$alcohol_category = cut(
#  wine_data$alcohol,
#  breaks = c(-Inf, 10, 12, Inf), # Define the thresholds for "Low", "Medium", "High"
#  labels = c("Low", "Medium", "High")
#)

# Convert to a Factor
#wine_data$alcohol_category = as.factor(wine_data$alcohol_category)

# View(wine_data)

```


## Data types


```{r}

str(wine_data)

```

```{r}
res_sugar_6580 = which(wine_data$residual.sugar == 65.80)
free_sulf_dio_289 = which(wine_data$free.sulfur.dioxide == 289.0)
dens_10103 = which(wine_data$density == 1.0103)

remove_idx = c(res_sugar_6580, free_sulf_dio_289, dens_10103)
wine_data = wine_data[-remove_idx, ]
nrow(wine_data)
```


## Fit additive Model

```{r}

model = lm(quality ~ ., data = wine_data)
summary(model)

```

## VIF Values


```{r}
if (!require(car)) install.packages("car")
library(car)

# Compute VIF Values
vif_values = vif(model)

# Display the VIF Values
vif_values


```

## Visualize correlations through a heatmap
```{r}
if (!require(corrplot)) install.packages("corrplot")
library(corrplot)
numeric_data = wine_data %>%
  select(-type, -quality) # Considers numeric predictors only
cor_matrix = cor(numeric_data, use = "complete.obs")


corrplot(cor_matrix, method = "color", addCoef.col = "black", 
         tl.cex = 0.8, number.cex = 0.7, main = "Correlation Heatmap")
```


## Fit interaction model

```{r}


model = lm(quality ~ .^2, data = wine_data)
summary(model)


```

## Including Plots


```{r}

# Scatter plot: Volatile Acidity vs Quality
plot(wine_data$volatile.acidity, wine_data$quality,
     main = "Volatile Acidity vs Wine Quality",
     xlab = "Volatile Acidity",
     ylab = "Wine Quality",
     col = "blue", pch = 19, cex = 0.5)

# Scatter plot: Residual Sugar vs Quality
plot(wine_data$residual.sugar, wine_data$quality,
     main = "Residual Sugar vs Wine Quality",
     xlab = "Residual Sugar",
     ylab = "Wine Quality",
     col = "green", pch = 19, cex = 0.5)


# Scatter plot: Fixed Acidity vs Quality
plot(wine_data$fixed.acidity, wine_data$quality,
     main = "Fixed Acidity vs Wine Quality",
     xlab = "Fixed Acidity",
     ylab = "Wine Quality",
     col = "purple", pch = 19, cex = 0.5)

# Scatter plot: Citric Acid vs Quality
plot(wine_data$citric.acid, wine_data$quality,
     main = "Citric Acid vs Wine Quality",
     xlab = "Citric Acid",
     ylab = "Wine Quality",
     col = "orange", pch = 19, cex = 0.5)

# Scatter plot: Chlorides vs Quality
plot(wine_data$chlorides, wine_data$quality,
     main = "Chlorides vs Wine Quality",
     xlab = "Chlorides",
     ylab = "Wine Quality",
     col = "red", pch = 19, cex = 0.5)

# Scatter plot: Free Sulfur Dioxide vs Quality
plot(wine_data$free.sulfur.dioxide, wine_data$quality,
     main = "Free Sulfur Dioxide vs Wine Quality",
     xlab = "Free Sulfur Dioxide",
     ylab = "Wine Quality",
     col = "cyan", pch = 19, cex = 0.5)

# Scatter plot: Total Sulfur Dioxide vs Quality
plot(wine_data$total.sulfur.dioxide, wine_data$quality,
     main = "Total Sulfur Dioxide vs Wine Quality",
     xlab = "Total Sulfur Dioxide",
     ylab = "Wine Quality",
     col = "brown", pch = 19, cex = 0.5)

# Scatter plot: Density vs Quality
plot(wine_data$density, wine_data$quality,
     main = "Density vs Wine Quality",
     xlab = "Density",
     ylab = "Wine Quality",
     col = "pink", pch = 19, cex = 0.5)

# Scatter plot: pH vs Quality
plot(wine_data$pH, wine_data$quality,
     main = "pH vs Wine Quality",
     xlab = "pH",
     ylab = "Wine Quality",
     col = "darkgreen", pch = 19, cex = 0.5)

# Scatter plot: Sulphates vs Quality
plot(wine_data$sulphates, wine_data$quality,
     main = "Sulphates vs Wine Quality",
     xlab = "Sulphates",
     ylab = "Wine Quality",
     col = "darkblue", pch = 19, cex = 0.5)

```


## Logs

```{r}
##model = lm(quality ~ .^2, data = wine_data)
##summary(model)


# Select only numeric columns
numeric_columns <- wine_data[sapply(wine_data, is.numeric)]

# Create histograms for each numeric column
par(mfrow = c(2, 2))  # Layout for multiple plots
sapply(names(numeric_columns), function(column) {
  hist(numeric_columns[[column]], main = paste("Histogram of", column),
       xlab = column, col = "lightblue", border = "white")
})

```


## Logs

```{r}
names(wine_data)
```


```{r}
model_add = lm(quality ~ ., data = wine_data)
model_log = lm(quality ~ fixed.acidity + log(volatile.acidity) + citric.acid + residual.sugar + chlorides + log(free.sulfur.dioxide) + total.sulfur.dioxide + density + (pH) + (sulphates) + (alcohol) + type, data = wine_data)

model_log_int = lm(quality ~ (fixed.acidity + log(volatile.acidity) + citric.acid + residual.sugar + chlorides + log(free.sulfur.dioxide) + total.sulfur.dioxide + density + (pH) + (sulphates) + (alcohol) + type)^2, data = wine_data)

summary(model_add)$adj.r.squared
summary(model_log)$adj.r.squared
summary(model)$adj.r.squared
summary(model_log_int)$adj.r.squared
```


```{r}
model_bac_aic = step(model_log_int, trace = 0)
model_bac_bic = step(model_log_int, k = log(nrow(wine_data)), trace = 0)
model_both_aic = step(model_log_int, direction = "both", trace = 0)
summary(model_bac_aic)$adj.r.squared
summary(model_bac_bic)$adj.r.squared
summary(model_both_aic)$adj.r.squared
```
```{r}
#model_bac_aic = step(model_add, trace = 0)
#model_bac_bic = step(model_add, k = log(nrow(wine_data)), trace = 0)
#model_both_aic = step(model_add, direction = "both", trace = 0)
#summary(model_bac_aic)$adj.r.squared
#summary(model_bac_bic)$adj.r.squared
#summary(model_both_aic)$adj.r.squared
```
